# Monte Carlo Method


Monte Carlo methods take advantage of the availability of:

  i. computer generated random variables 
  ii. the law of large numbers
  iii. the central limit theorem


**Notions:** 

* **Monte Carlo Method**: The use of (pseudo) random sampling as a tool to produce observations on which statistical inference can be performed to extract information about a system.
* **Monte Carlo Integration**: A specific Monte Carlo method, where we randomly sample uniformly over some domain and use the produced sample to construct an estimator for a **univariate and multivariate  integral**. (Integrals are everywhere in statistics!)
* **Stochastic Simulation** (or **Monte Carlo Simulation**): The application of the Monte Carlo method.  


\


## Classical Monte Carlo Integration


The generic problem here is the evaluation of integrals. (Be aware: Integrals are everywhere in statistics!). For instance, 
$$
\mathbb{E}_f\left(h(X)\right)=\mathbb{E}\left(h(X)\right)=\int_\mathcal{X}h(x)\,f(x)\,dx.
$$


**Convergence:**

Given our previous developments, it is natural to propose using a realization $x_1,\dots,x_m$ from a (pseudo random) i.i.d. sample $X_1,\dots,X_m$ with each $X_j$ distributed as $X\sim f$ to approximate the above integral by the empirical mean
$$
\bar{h}_m=\frac{1}{m}\sum_{j=1}^m h(x_j).
$$
By the [Strong Law of Large Numbers](http://www.statlect.com/asylln1.htm) we know that the empirical mean $\bar{h}_m$ converges almost surely (a.s.) to the desired limit $\mathbb{E}\left(h(X)\right)$ as $m\to\infty$. (The only prerequisits are that $f$ has finite first moments, i.e., $\mathbb{E}\left(h(X)\right)<\infty$, and that $\bar{h}_m$ is constructed from an i.i.d. sample $X_1,\dots,X_m$.)

As we can use the computer to produce realizations from the i.i.d. sample $X_1,\dots,X_m$, we can in principle choose an arbitrary **large sample** size $m$ such that $\bar{h}_m$ can (in principle) be **arbitrarily close** to the desired limit $\mathbb{E}\left(h(X)\right)$. 


**Though,** ... 

* ... which sample size $m$ is large enough? 
* Or "equivalently": How fast converges $\bar{h}_m$ to the desired limit $\mathbb{E}\left(h(X)\right)$?

\

**Speed of Convergence:**

OK, we know now that $\bar{h}_m$ reaches its limit (here in the "almost surely" sense) as $m\to\infty$ under some rather loose conditions on the random sample $X_1,\dots,X_m$. 

If we are willing to additionally assume that $f$ has finite second moments, i.e., $\mathbb{E}(h(X)^2)<\infty$, we can additionally say something about **how fast** $\bar{h}_m$ converges (a.s.) to $\mathbb{E}(h(X))$. 

The **speed of convergence** of the stochastic sequence $\{\bar{h}_m\}$ (i.e., now we think of $\bar{h}_m$ as the {\color{red}RV} $\bar{h}_m=\frac{1}{m}\sum_{j=1}^m h({\color{red}{X_{j}}})$) to its limit $\mathbb{E}(h(X))$ can be assessed by answering the question how fast the standard deviation (which is a function of $m$) of the stochastic sequence converges to zero as $m\to\infty$. 

* The variance of $\bar{h}_m$ is given by 
  $$
  \mathbb{V}_f\left(\bar{h}_m\right)=
  \mathbb{V}\left(\frac{1}{m}\sum_{j=1}^m h(X_j)\right)=
  \frac{1}{m}\mathbb{V}\left(h(X)\right)
  $$

* Note that assuming finite second moments $\mathbb{E}(h(X)^2)<\infty$ is equivalent to assuming finite variance $\mathbb{V}\left(h(X)\right)<\infty$. Consequently, we can set   $\mathtt{const}=\sqrt{\mathbb{V}\left(h(X)\right)}$ with $0<\mathtt{const}<\infty$ such that
  $$
  \sqrt{\mathbb{V}\left(\bar{h}_m\right)}=m^{-1/2}\mathtt{const}\propto m^{-1/2}.
  $$

I.e., the speed of convergence (or rate) of the stochastic sequence $\{\bar{h}_m\}$ is proportional to the deterministic sequence $\{m^{-1/2}\}$.   

\


**Remark:**
Even if we would not know the value of $\mathtt{const}=\sqrt{\mathbb{V}\left(h(X)\right)}$, we know now that the improvement from $m=10$ to $m=100$ will be *much* higher than from $m=110$ to $m=200$. In practice, a typical choice is $m=10000$; for moderate standard deviations this choice will guarantee a very good approximation.


\


**Limit Distribution:**

Of course, we can estimate the variance of the estimator $\mathbb{V}\left(\bar{h}_m\right)$ by its empirical version
$$
v_m=\frac{1}{m}\left(\frac{1}{m}\sum_{j=1}^m\left(h(x_j)-\bar{h}_m\right)^2\right),
$$
where again by the [Strong Law of Large Numbers (SLLN)](http://www.statlect.com/asylln1.htm)
$$
\left(\frac{1}{m}\sum_{j=1}^m\left(h(x_j)-\bar{h}_m\right)^2\right)\to_{\text{a.s.}}\mathbb{V}\left(h(X)\right).
$$
<!--
I.e., after some rewriting, we have that: 
$$
\frac{\bar{h}_m - \mathbb{E}\left(h(X)\right)}{\sqrt{v_m}}\to_{\text{a.s.}}
\sqrt{m}\left(\frac{\bar{h}_m - \mathbb{E}\left(h(X)\right)}{\sqrt{\mathbb{V}\left(h(X)\right)}}\right).
$$
-->
By the [Central Limit Theorem (CLT)](http://www.statlect.com/central_limit_theorem.htm) we have 
$$
\sqrt{m}\left(\frac{\bar{h}_m - \mathbb{E}\left(h(X)\right)}{\sqrt{\mathbb{V}\left(h(X)\right)}}\right)\to_d Z,
$$
where $Z\sim N(0,1)$. Note that the the above sequence $\{\sqrt{m}\}$ **just hinders** the convergence of the sequence $\bar{h}_m - \mathbb{E}\left(h(X)\right)\to_{a.s.}0$ such that the quotient converges to a "stable" distribution. 


The above result can now be used for the construction of (asymptotically valid) **convergence tests** and **confidence intervals** with respect to $\bar{h}_m$, since for large $m$ 
$$
\bar{h}_m\,\overset{d}{\approx} N\left(\mathbb{E}\left(h(X)\right),\frac{\mathbb{V}\left(h(X)\right)}{m}\right).
$$

And as we can use the computer to generate realizations of the i.i.d. sample $X_1,\dots,X_m$ from a generic $X\sim f$, we can easily approximate the mean $\mathbb{E}\left(h(X)\right)$ and the variance $\mathbb{V}\left(h(X)\right)$ with arbitrary accuracy as $m\to\infty$; by the SLLN (or the WLLN).


\


**Example: A first Monte Carlo Integration**

Let's say we want to integrate the function $h(x)=\left(\cos(50\,x)+\sin(20\,x)\right)^2$. Although this function could be integrated analytically it is a good first test case. The left plot below shows the graph of the function $h(.)$.

To approximate the integral 
$$
\int_\mathcal{X}h(x)dx\quad\text{with}\quad\mathcal{X}=[0,1]
$$ 
we can use that
$$
\int_\mathcal{X}h(x)dx=\int_\mathcal{[0,1]}1\cdot h(x)dx =\mathbb{E}_{f_\text{Unif[0,1]}}(h(X)).
$$ 

Thus, we generate a realization $(u_1,\dots,u_n)$ from the i.i.d. random sample $U_1,\dots,U_n\sim[0,1]$ and approximate 
$$
\int_\mathcal{X}h(x)dx\approx \bar{h}_n=\frac{1}{n}\sum_{i=1}^n h(u_i).
$$ 

In order to assess how good this approximation is, we need to consider the stochastic propoerties of the RV 
$$
\frac{1}{n}\sum_{i=1}^n h(U_i).
$$ 
This is done using the above (review of) results on the limit distribution of the sample mean which allows us to construct an approximative $95\%$ confidence interval, since for large $n$ 
$$
\left[\bar{h}_n - 1.96\frac{\mathtt{std.error}_n}{\sqrt{n}}, \bar{h}_n + 1.96\frac{\mathtt{std.error}_n}{\sqrt{n}}\right]\approx
\left[\bar{h}_n - 1.96  \sqrt{\frac{\mathbb{V}(h(U_i))}{n}}, \bar{h}_n + 1.96  \sqrt{\frac{\mathbb{V}(h(U_i))}{n}}\right],
$$
where $\mathtt{std.error}_n^2=n^{-1}\sum_{i=1}^n(h(u_i)-\bar{h}_n)^2$. 

The right plot below shows one realization of the stochastic sequence $\{\bar{h}_1,\dots,\bar{h}_n\}$ with $n=10000$, where the realized value of $\bar{h}_n$ is $0.966$. This compares favorably with the with the exact value of $0.965$.



```{r,echo=FALSE, out.width='\\textwidth', fig.align='center', fig.width=10, fig.height=5}
library("scales")
# h(x):
h_fun <- function(x){
  result <- (cos(50*x)+sin(20*x))^2
  return(result)
}

# sample size
n <- 10000

# Generate sample of uniforms
set.seed(321)
u_vec <- runif(n=n)

# Approximation of the integral 
# More precisely: This is one realization of the stochastic 
# sequence \bar{h}_1,...,\bar{h}_n.
h_bar_n <- cumsum(h_fun(u_vec))/c(1:n)

# 'Final' value: 
# round(h_bar_n[n], digits = 3)
# True value:
true.value <- 0.965

# 95% CI
# Standard error of the estimator using the algebraic 
# formula for the variance (german: 'verschiebungssatz')
st.error_n <-  sqrt((cumsum(h_fun(u_vec)^2) - cumsum(h_fun(u_vec))^2/(1:n))/(1:n))
CI_u       <-  h_bar_n + 1.96 * st.error_n / sqrt(1:n)
CI_l       <-  h_bar_n - 1.96 * st.error_n / sqrt(1:n)

par(mfrow=c(1,2))
# plot 1:
xx  <- seq(from=0, to=1, len=500)
plot(x = xx, y = h_fun(xx), type="l", main="Function h(.)", xlab="x", ylab="h(x)")
# plot 2: 
plot(x = c(1:n), y = h_bar_n, type="n", ylim=c(0.7,1.2), 
     xlab = "n", ylab = "")
polygon(x = c(1:n, rev(1:n)), y = c(CI_u, rev(CI_l)), 
        col=alpha("blue", 0.5), border=alpha("blue", 0.5))
lines(x = c(1:n), y = h_bar_n, type="l")
lines(x = c(1:n), y = rep(true.value, n), type="l", col="red")
##
legend("topright", legend = c(expression(bar(h)[n]), "True Value", "95% CI"), lty=c(1,1,0), pch=c(22,22,22), pt.cex=c(0,0,2),
     pt.bg = c("black", "red", alpha("blue", 0.5)), 
     col =   c("black", "red", alpha("blue", 0.5)))
```
```{r, echo=FALSE}
par(mfrow=c(1,1))
```

\

**Remarks:**

* The approach followed in the above example can be successfully utilized in many cases, even though it is often possible to achieve greater efficiency through numerical methods (e.g., [Riemann Sum](https://en.wikipedia.org/wiki/Riemann_sum), [Trapezoidal Rule](https://en.wikipedia.org/wiki/Trapezoidal_rule), [Simpson's Rule](https://en.wikipedia.org/wiki/Simpson%27s_rule), etc.) in dimensions 1 or 2. 
* The approach is particularly useful for approximating integrals over higher dimensional sets. 


\


**Example: Approximation of Normal Distribution Tables**

A possible way to construct normal distribution tables is to use MC simulations. 

Generate a realization $(x_1,\dots,x_n)$ from an i.i.d. standard normal random sample, e.g., using the Box-Muller algorithm. 

The approximation of the standard normal cdf
$$
\Phi(t)=\int_{-\infty}^t\frac{1}{\sqrt{2\pi}}e^{-y^2/2}dy
$$
by the Monte Carlo method is thus
$$
\hat{\Phi}_n(t)=\frac{1}{n}\sum_{i=1}^n 1_{(x_i\leq t)}.
$$
The corresponding RV $\hat{\Phi}_n(t)=\frac{1}{n}\sum_{i=1}^n1_{(X_i\leq t)}$ has (exact) variance 
$$
\mathbb{V}(\hat{\Phi}_n(t))=\frac{\Phi(t)(1-\Phi(t))}{n},
$$
since the single RVs $1_{(X_i\leq t)}$ are independent Bernoulli with success probability $\Phi(t)$. 

For values of $t$ around $t=0$, the variance is thus approximately $1/4n$. 

To achieve a precision of **four decimals** by means of a $99.9\%$ confidence interval, the approximation requires on average $n\approx 10^8$ simulations. 

The table below gives the evolution of this approximation for several values of $t$ and shows a very accurate evaluation for $n=10^8$. 

\

$$
\begin{array}{cccccccccc}
\hline
n   &t=0.0  &t=0.67 &t=0.84 &t=1.28 &t=1.65 &t=2.32 &t=2.58 &t=3.09 &t=3.72 \\
\hline
10^2 &0.485  &0.74   &0.77   &0.9    &0.945  &0.985  &0.995  &1      &1      \\
10^3 &0.4925 &0.7455 &0.801  &0.902  &0.9425 &0.9885 &0.9955 &0.9985 &1      \\
10^4 &0.4962 &0.7425 &0.7941 &0.9    &0.9498 &0.9896 &0.995  &0.999  &0.9999 \\
10^5 &0.4995 &0.7489 &0.7993 &0.9003 &0.9498 &0.9898 &0.995  &0.9989 &0.9999 \\
10^6 &0.5001 &0.7497 &0.8    &0.9002 &0.9502 &0.99   &0.995  &0.999  &0.9999 \\
10^7 &0.5002 &0.7499 &0.8    &0.9001 &0.9501 &0.99   &0.995  &0.999  &0.9999 \\
10^8 &0.5    &0.75   &0.8    &0.9    &0.95   &0.99   &0.995  &0.999  &0.9999 \\
\end{array}
$$

\

**Remarks:**

* To achieve a precision of **two decimals** by means of a $99.9\%$  confidence interval, already $n=10^4$ leads to satisfactory results. 
* Note that **greater accuracy is achieved in the tails** and that more efficient simulation methods could be used (e.g., Importance Sampling). 

\


## Importance Sampling

Importance sampling aims to reduce the variance of the Monte Carlo integral estimate therefore it's refereed to as a **variance reduction** technique. This variance reduction is achieved by weighting functions, so-called **importance functions**. 

As in the case of Monte Carlo integration the focus lies on evaluating the integral
$$
\mathbb{E}_f(h(X))=\int_\mathcal{X}h(x)f(x)\,dx.
$$

* Though, it turns out that the above approach, i.e., sampling from $f$ is often suboptimal. 

* Observe that the value of the above integral can be represented by infinitely many alternative choices of the triplet $(\mathcal{X}, h, f)$. Therefore, the search for an optimal estimator should encompass all these possible representations.

Let's illustrate this with a simple example.


**Example: Cauchy Tail Probability (from [Ripley 1987](http://onlinelibrary.wiley.com/book/10.1002/9780470316726))**

Suppose that the quantity of interest is the probability, say $p$, that a [Cauchy](https://en.wikipedia.org/wiki/Cauchy_distribution) $\mathrm{C}(0,1)$ RV is larger than $2$, i.e.:
$$
p=\int_{2}^{+\infty}\frac{1}{\pi(1+x^2)}\,dx.
$$

**1. Naive Approach:** 
If $p$ is approximated through the empirical mean
$$
\hat{p}_{1}=\frac{1}{m}\sum_{j=1}^m1_{(X_j>2)}
$$
of an i.i.d. sample $X_1,\dots,X_m\sim\mathrm{C}(0,1)$, then the variance of this estimator, a [binomial](https://en.wikipedia.org/wiki/Binomial_distribution) RV scaled by $1/m$, is 
$$
\mathbb{V}(\hat{p}_{1})=\frac{1}{m^2}\mathbb{V}\left(\sum_{j=1}^m1_{(X_j>2)}\right)=\frac{p(1-p)}{m},
$$
which is equal to $0.1275/m$, since (we already know that) $p=0.15$. 

\



**2. Accounting for Symmetry (i.e., using the 'Adjusting Screws' $\mathcal{X}$ and $h$):**
We can achieve a **more efficient estimator** (i.e., an estimator with lower variance for a given same sample size $n$) if we take into account the symmetric nature of $\mathrm{C}(0,1)$. Obviously, our target integral can be equivalently written as
$$
p=\frac{1}{2}\left(\int_{-\infty}^{-2}\frac{1}{\pi(1+x^2)}\,dx + \int_{2}^{+\infty}\frac{1}{\pi(1+x^2)}\,dx \right).
$$
This representation has the attractive feature that we can use a much higher fraction of the simulated data by using the following new empirical mean:
$$
\hat{p}_{2}=
\frac{1}{2}\left(\frac{1}{m}\sum_{j=1}^m1_{(X_j<-2)}+ \frac{1}{m}\sum_{j=1}^m1_{(X_j>2)}\right)\;=\;
\frac{1}{2m}\sum_{j=1}^m1_{(|X_i|>2)}.
$$
The variance of this new estimator,
$$
\mathbb{V}(\hat{p}_{2})=\frac{1}{4m^2}\mathbb{V}\left(\sum_{j=1}^m1_{(|X_i|>2)}\right)=\frac{2p(1-2p)}{4m},
$$
is equal to $0.0525/m$, i.e., lower than in the naive approach.

\



**3. Using all 'Adjusting Screws' $\mathcal{X}$, $h$, and $f$:**
The (relative) inefficiency of the above approaches is due to the generation of RVs outside the domain of interest, $[2,+\infty)$, which are in some sense irrelevant for the approximation of $p$. This motivates the following reformulation of $p$: 

By symmetry of $f$:
$$
\frac{1}{2}=\int_{0}^2\frac{1}{\pi(1+x^2)}dx + \underbrace{\int_{2}^{+\infty}\frac{1}{\pi(1+x^2)}dx}_{=p}
$$
$$
\Leftrightarrow \; p=\frac{1}{2}-\int_{0}^2\frac{1}{\pi(1+x^2)}dx.
$$
Furthermore, we can re-arrange the last integral a bit such that 
$$
\int_{0}^2\;\left(\frac{1}{2}\cdot 2\right)\;\frac{1}{\pi(1+x^2)}\,dx = 
\int_{0}^2\;\underbrace{\frac{1}{2}}_{f_{\mathrm{Unif}[0,2]}}\;\underbrace{\frac{2}{\pi(1+x^2)}}_{=h(x)}\,dx = 
\mathbb{E}(h(U)),\quad\text{where}\quad U\sim\mathrm{Unif}[0,2].
$$


Therefore a new alternative method for evaluating $p$ is:
$$
\hat{p}_{3}=\frac{1}{2} - \frac{1}{m}\sum_{j=1}^m h(U_j),\quad\text{where}\quad U_j\sim\mathrm{Unif}[0,2].
$$
Using integration by parts, it can be shown that $\mathbb{V}(\hat p_3)=0.0285/m$. (Compare this to the former results: $\mathbb{V}(\hat{p}_{2})=0.0525/m$ and $\mathbb{V}(\hat{p}_{1})=0.1275/m$.)

\


**A More General Point of View:**

The idea of importance sampling is related to weighted and stratified sampling ideas. As illustrated by the above example, when estimating 
$$
\theta=\mathbb{E}_f(h(X))=\int h(x)f(x)dx.
$$

Some outcomes of $X\sim f$ may be more important than others in determining $\theta$ and we wish to select such values more frequently. 

For instance, if $\theta$ denotes the probability of the occurrence of a very rare event, then the only way to estimate $\theta$ at all accurately may be to produce the rare events more frequently. 

To achieve this, we can simulate a model which gives pdf $g$ to $X$ instead of the correct pdf $f$, where both pdfs need to be known. This can be easily done, since
$$
\theta=\mathbb{E}_f(h(X))=\int h(x)\left(\frac{g(x)}{g(x)}\right)\;f(x)dx=
\int \underbrace{\left(h(x)\frac{f(x)}{g(x)}\right)}_{=\psi(x)}\;g(x)dx=
\int \psi(x)\;g(x)dx=
\mathbb{E}_g(\psi(X)).
$$

This leads to the following unbiased estimator for $\theta$ based on sampling from $g$:
$$
\hat{\theta}_g=\frac{1}{n}\sum_{i=1}^n\psi(X_i)\quad\text{with}\quad X_i\sim g,
$$
which is a weighted mean of the $h(X_i)$ with weights inversely proportional to the "**selection factor**" $\frac{g(X_i)}{f(X_i)}$.
<!-- 
For appropriately chosen pdfs f and g: 
The *selection factor* refers to "how more likely is it to select a 'rare event'?".
The inverse weight re-scales these 'too often' chosen rare events. 
-->


For the variance of the estimator $\hat{\theta}_g$ we have
$$
\mathbb{V}(\hat{\theta}_g)=\frac{1}{n}\mathbb{V}(\psi(X_i))=
\frac{1}{n}\int\left(\psi(x)-\theta\right)^2g(x)dx=
\frac{1}{n}\int\left(\frac{h(x)\,f(x)}{g(x)}-\theta\right)^2g(x)dx,
$$
which, depending on the choice of $g(.)$, can be much smaller (or larger) than the variance of the naive estimator from the classical Monte Carlo Integration using the ordinary empricial mean.
<!-- 
$$
\mathbb{V}(\hat{\theta}_{\text{naive}})=
\mathbb{V}\left(\frac{1}{n}\sum_{i=1}^n 1_{(X_i\in\{\mathtt{Rare.Event}\})}\right)=
\frac{1}{n}\mathbb{V}\left(1_{(X_i\in\{\mathtt{Rare.Event}\})}\right)=\frac{\theta(1-\theta)}{n}.
$$ 
-->

\
\


*Minimum Variance Theorem*

:    The **importance function** $g(.)$ which minimizes the variance $\mathbb{V}(\psi(X_i))$, and therefore the variance $\mathbb{V}(\hat{\theta}_g)$,      is given by 
     $$
     g^\ast(x)=\frac{|h(x)|f(x)}{\int |h(z)|f(z)dz}.
     $$ 


**Proof:** Done in the lecture. 

\


Though, this result is rather formal (in the sense of "impractical"), since, e.g., if $h(x)>0$ then $g^\ast$ requires us to know $\int h(z)f(z)dz$, which is just the integral of interest! 

**Remarks:**

The above minimum variance result is still useful:

*  It tells us that a good choice of $g(x)$ shall mimic the shape of $|h(x)|f(x)$, since the optimal $g^\ast(x)\propto |h(x)|f(x)$.
*  Furthermore, $g(x)$ should be chosen such that it has a thicker tail than $f(x)$, since the variance $\mathbb{V}(\hat{\theta}_g)$ crucially depends on the quotient $f(x)/g(x)$ which would "explode" for $g(x)\approx 0$.


\


Let's apply our new insights to the above example on the Cauchy tail probability $p$. 

**Example: Cauchy Tail Probability (cont.)**


Above we had: 

1. $f(x)=\frac{1}{\pi(1+x^2)}$, the pdf of $\mathrm{C}(0,1)$ and
2. $h(x)=1_{(x>2)}$, i.e., here $|h(x)|=h(x)$.

Therefore
$$
p=\mathbb{E}_f(h(X))=\int h(x)f(x)dx=\int_{2}^{\infty}f(x)dx=\int_{2}^{\infty}\underbrace{\frac{f(x)}{g(x)}}_{=\psi(x)}\;g(x)dx=\mathbb{E}_g(\psi(X)),
$$
where the $h$ function is absorbed by the formulation of the definite integral.

A possibly good (and simple) choice of $g$ is, e.g., $g(x)=2/(x^2)$, since this function:

* "closely" matches $h(x)f(x)$ and 
* $g$ has thicker tails than $f$. 


```{r, fig.width=7, fig.height=4, fig.align='center', echo=FALSE}
# g(x):
g_fun <- function(x){2/(x^2)}
# 
xx    <- seq(from= 0, to=15, length.out = 500)
# plot
plot(x = xx, dcauchy(xx, location = 0, scale = 1), 
     type="l", xlab="", ylab="", ylim=c(0,0.55), axes = FALSE)
axis(side = 1, at=c(0,2,5,10,15)); axis(side = 2)
# add graph of g(x)
lines(x = xx[xx>2], g_fun(xx[xx>2]), col="red")
# region of interest selected by h(x): x>2
abline(v=2, lty=2)
# legend
legend("topright", legend = c("pdf-Cauchy(0,1)", expression(g(x)==2/(x^2))), lty=c(1,1),
     col = c("black", "red"))
```

\

**Caution:** It is not straight forward to directly sample from $g$, therefore we need some further steps:

\

The choice of $g$ leads to 
$$
p=\mathbb{E}_g(\psi(X))=
\int_{2}^{+\infty}\left(\frac{x^2}{2\,\pi(1+x^2)}\right)\,\frac{2}{x^2}\,dx=
\int_{2}^{+\infty}\left(\frac{1}{\pi(1+x^{-2})}\right)\,x^{-2}\,dx.
$$

\


Now we can apply some additional (rather case-specific) re-arrangements:

Integration by substitution (substituting $u=x^{-1}$) yields:
$$
p=\int_{0}^{1/2}\frac{1}{\pi(1+u^2)}du.
$$
Again, we can re-arrange the last integral a bit such that
$$
p=\int_{0}^{1/2}\underbrace{2}_{f_{\mathrm{Unif}[0,1/2]}}\;\underbrace{\frac{1}{2\,\pi(1+u^2)}}_{=h(u)}\,du=\mathbb{E}(h(U)),\quad\text{where}\quad U\sim\mathrm{Unif}[0,1/2].
$$
Therefore, we have a final fourth version of the estimator of $p$:
$$
\hat{p}_4=\sum_{j=1}^m h(U_j),\quad\text{where}\quad U\sim\mathrm{Unif}[0,1/2] 
$$
and $h(u)=1/(2\pi(1+u^2))$. 

The variance of $\hat{p}_4$ is $(\mathbb{E}(h(U)^2)-\mathbb{E}(h(U))^2)/m$ and an integration by parts shows that $\mathbb{V}(\hat{p}_4)=0.95\cdot 10^{-4}/m$. Compare this to the former results: $\mathbb{V}(\hat p_3)=0.0285/m$, $\mathbb{V}(\hat{p}_{2})=0.0525/m$ and $\mathbb{V}(\hat{p}_{1})=0.1275/m$. The variance of $\hat{p}_4$ is by a factor of $10^{-3}$ lower than the variance of the original $\hat{p}_1$.


<!--
**Version 2:** It can be shown that (see [Ripley 1987](http://onlinelibrary.wiley.com/book/10.1002/9780470316726))
$$
p=\int_0^{1/2}\frac{y^{-2}}{\pi(1+y^{-2})}dy,
$$
where this integral can also be seen as the expectation of 
$$
\frac{1}{4}h(U)=\frac{1}{2\pi(1+U^2)},\quad\text{where}\quad U\sim\mathrm{Unif}[0,1/2].
$$
-->




<!--
## The 'Real Need' of Monte Carlo Simulation 

**Monte Carlo Simulation applied to Hypothesis Tests**

* **Problem:** Very often, statistical test procedures (or estimators) rely on asymptotic arguments. Asymptotic arguments ease the live of a statistician. In practice, however, we never have something like a diverging sample size of $n\to\infty$, but need to deal with a finite sample size $n$. All we can hope for is that the asymptotic results (e.g., on the level of significance of a test statistic and its power) are good approximations to the finite $n$ case. 

* **Solution:** Monte Carlo simulation the classical tool to investigate the finite $n$ performance of statistical test procedures (or estimators). It helps to answer questions like: How good are the asymptotic results given finite sample size scenarios of $n=100$, $n=500$, etc.

For instance, the **likelihood ratio (LR)** test statistic  
$$
-2\,\log\left[\ell(\hat\theta|x)/\ell(\hat\theta_0|x)\right]=-2\,\left\{\log\ell(\hat\theta|x)-\log\ell(\hat\theta_0|x)\right\}\to_d\chi^2_r
$$
is distributed as $\chi^2_r$ generally only in the limiting case as $n\to\infty$; and under some regularity constraints on the likelihood function. In the formula above $\ell(\theta|x)$ denotes the likelihood function, $\hat\theta\in\mathbb{R}^k$ is the estimated (via maximum likelihood) parameter vector from the unconstrained model and $\hat\theta_0\in\mathbb{R}^k$ is the estimated parameter vector from the constrained model with $r\leq k$ restrictions.

\

**Example: Contingency Tables**

The following table gives the results of a study comparing radiation therapy with surgery in treating cancer of the larynx. 


$$
\begin{array}{c|cc|c}
                   &  \text{Cancer}  &  \text{Cancer not}&           \\
                   &  \text{Controlled}  &  \text{Controlled}&           \\
\hline                   
\text{Surgery}     &  y_{11}=21      &   y_{12}=2        &  n_{1.}=23\\
\text{Radiation}   &  y_{21}=15      &   y_{22}=3        &  n_{2.}=18\\
\hline
                   &  n_{.1}=36      &   n_{.2}=5        &  n=41
\end{array}
$$

\

Let's condition on a fix number of total observations $n$. Then the random vector $(Y_{11},Y_{12},Y_{21},Y_{22})^\top$ comes from a [multinomial distribiton](https://en.wikipedia.org/wiki/Multinomial_distribution) with 4 cells and cell probabilities 
$$
p=(p_{11}, p_{12}, p_{21}, p_{22}),\quad\text{with}\quad\sum_{ij}p_{ij}=1,
$$
that is, 
$$
(Y_{11},Y_{12},Y_{21},Y_{22})^\top\sim\mathcal{M}_4(n, p).
$$
With $y_{ij}$ denoting the number of realizations in cell $ij$, the likelihood function can be written as 
$$
\ell(p|y)\propto\prod_{ij}p_{ij}^{y_{ij}},
$$
where the 4-dimensional parameter space can be displayed as following:
$$
\begin{array}{cc|c}
p_{11} & p_{12} & p_{1.}\\
p_{21} & p_{22} & p_{2.}\\
\hline
p_{.1} & p_{.2} & 1\\
\end{array}
$$



**Null Hypothesis:** The null hypothesis to be tested is one of independence, which is to say that the surgery treatment has no bearing on the control of cancer. Translated into a parameter statement this means
$$\text{H}_0: p_{11}=p_{1.}\,p_{.1}\quad\text{against}\quad\text{H}_1: p_{11}\neq p_{1.}\,p_{.1}.$$


The likelihood ratio statistic for testing this hypothesis is 
$$
\lambda=\frac{\max_{p\text{ s.t. }p_{11}=p_{1.}p_{.1}}\ell(p|y)}{\max_{p}\ell(p|y)}.
$$
It is "straightforward" to show that the denominator maximum is attained at:
$$
\hat{p}_{ij}=\frac{n_{ij}}{n}\quad\text{for all}\quad ij. 
$$
and the numerator maximum at
$$
\hat{p}_{11}=\hat{p}_{1.}\hat{p}_{.1}\quad\text{with}\quad \hat{p}_{1.}=\frac{n_{1.}}{n}\quad\text{and}\quad\hat{p}_{2.}=\frac{n_{2.}}{n},
$$
$$
\text{and} \quad \hat{p}_{ij}=\frac{n_{ij}}{n}\quad\text{for all other}\quad ij. 
$$

As mentioned above, under H$_0$, $-2\log \lambda$ is asymptotically distributed as $\chi^2_1$. However, with only $n=42$ observations, the asymptotics do not necessarily apply. One alternative is to use devise a **Monte Carlo experiment** to simulate the null distribution of $-2\log \lambda$ or equivalently of $\lambda$ in order to obtain a cutoff point for a hypothesis test. A more sophisticated approach is that of [Mehta at el. (2000)](http://amstat.tandfonline.com/doi/abs/10.1080/01621459.2000.10473906).  

**Description of the Procedure:**

* Let's denote the finite $n$ null distribution of $\lambda$ by $f_{0,n}(.)$. As we are interested in an $\alpha$ level test, we need to specify $\alpha$ (e.g., $\alpha=0.05$) and to solve the following integral for the $1-\alpha$ quantile $\lambda_\alpha$:
$$
\int_0^{\lambda_\alpha}f_{0,n}(u)du=1-\alpha.
$$

* The standard Monte Carlo approach to this problem is to generate random variables $\lambda_k\sim f_{0,n}$, $k=1,\dots,m$, then order the sample 
$$
\lambda_{(1)}\leq \lambda_{(2)}\leq\dots\leq \lambda_{(m)}
$$
and finally calculate the empirical $1-\alpha$ quantile $\lambda_{(\lfloor (1-\alpha)\,m\rfloor)}$, where $\lfloor x \rfloor$ is the next lower integer to $x$, e.g., $\lfloor 2.9 \rfloor=2$.

* Similarly to the above integration example which builds on the SLLN, the central idea here is to use the so-to-say SLLN for quantiles:
$$
\lambda_{(\lfloor (1-\alpha)\,m\rfloor)}\to_{a.s.} \lambda_{\alpha}\quad\text{as}\quad m \to\infty;
$$
see, e.g., the classical book "Approximation Theorems of Mathematical Statistics" of R. Serfling Ch. 2.3.1. As a computer is doing this job for us, we can in principle choose an arbitrary large $m$ such that the above approximation of $\lambda_{\alpha}$ by $\lambda_{(\lfloor (1-\alpha)\,m\rfloor)}$ can be arbitrarily good. 

To run the 

```{r}
set.seed(123)
#
p_init <- runif(4)
#
p_11   <- p_init[1]
p_12   <- p_init[2]
p_21   <- p_init[3]
p_22   <- p_init[4]
#
p_1.   <- p_11 + p_12
p_.1   <- p_11 + p_21
#
p_11   <- p_1. * p_.1

# probabilities under H0:
p_0 <- c(p_11, p_21, p_12, p_22)


Y_vec <- rmultinom(n=1, size=4, prob=p_0)

```
-->

<!-- 
**Example: LR Test**

Let's assume we want to test the following regression model 
$$
Y_t=\beta_0 + \beta_1 t +\varepsilon_{t}
$$
against the constrained model without a time trend:
$$
Y_t=\beta_0 + \varepsilon_{t},
$$
where $\varepsilon_{t}\sim N(0,\sigma_\varepsilon^2)$ with $0<\varepsilon^2<\infty$.


```{r}
LL_fun <- function(beta0, beta1, sigma) {
      # residuals
      res_vec   <-  y - x * beta1 - beta0
      # log-transformed probabilities of observing the vector of residuals res_vec: 
      Log_probs <- dnorm(x = R, mean = 0, sd = sigma, log = TRUE)
      # 
      result <- - sum(Log_probs)
      return(result)
}
```
-->


