## Solutions {-}

1. Deriving the upper bound in @exm-AREfficiency ("Normals from Double Exponentials"): 
$$
\begin{align*}
\frac{f(x)}{g(x|b)} 
&= \frac{\frac{1}{\sqrt{2\pi}}\exp\left(-\frac{x^2}{2}\right)}{\frac{1}{2b}\exp\left(-\frac{|x|}{b}\right)}\\[2ex]
&= \sqrt{\frac{2}{\pi}}\;b\; \exp\left(-\frac{x^2}{2}\right) \exp\left(\frac{|x|}{b}\right)\\[2ex]
&= \sqrt{\frac{2}{\pi}}\;b\; \exp\left(\frac{2|x| - x^2b}{2b}\right)\\[2ex]
&=\left\{
 \begin{array}{ll}
 \sqrt{\frac{2}{\pi}}\;b\; \exp\left(\frac{2 x  - x^2b}{2b}\right) &\;\text{for}\; x\geq 0\\
 \sqrt{\frac{2}{\pi}}\;b\; \exp\left(\frac{-2 x  - x^2b}{2b}\right) &\;\text{for}\; x <0\\
 \end{array}
\right.
\end{align*}
$$
Consider a given value of $b>0,$ and the case of $x\geq 0.$ Maximizing  
$$
\begin{align*}
\exp\left(\frac{2x - x^2b}{2b}\right) 
& = \exp\left(\frac{1}{b}x - \frac{1}{2}x^2 \right)
\end{align*}
$$ 
maximal with respect to $x\geq 0$ yields
$$
\begin{align*}
\exp\left(\frac{1}{b}x - \frac{1}{2}x^2 \right) \cdot \left(\frac{1}{b} - x \right) &\overset{!}{=} 0
\Leftrightarrow 
x=\frac{1}{b}
\end{align*}
$$ 
Correspondingly, for a given value of $b>0,$ and the case of $x<0,$ maximizing 
$$
\begin{align*}
\exp\left(\frac{-2x - x^2b}{2b}\right) 
& = \exp\left(-\frac{1}{b}x - \frac{1}{2}x^2 \right)
\end{align*}
$$ 
yields
$$
\begin{align*}
\exp\left(-\frac{1}{b}x - \frac{1}{2}x^2 \right) \cdot \left(-\frac{1}{b} - x \right) &\overset{!}{=} 0
\Leftrightarrow 
x=-\frac{1}{b}
\end{align*}
$$ 

```{r}
b <- 2
myfun <- function(x, b){exp( (2*abs(x) - x^2 * b)/(2 * b) )}
xx <- seq(-10, 10, len = 500)
yy <- myfun(xx, b = b)
plot(x = xx, y = yy, type = "l")
abline(v=c(-1/b,1/b), lty=2)
```
Thus
$$
\begin{align*}
\frac{f(x)}{g(x|b)} 
& \leq 
\left\{
 \begin{array}{ll}
 \sqrt{\frac{2}{\pi}}\;b\; \exp\left(\frac{2 \frac{1}{b}  - \frac{1}{b^2} b}{2b}\right) &\;\text{for}\; x\geq 0\\
 \sqrt{\frac{2}{\pi}}\;b\; \exp\left(\frac{2 \frac{1}{b}  - \frac{1}{b^2} b}{2b}\right) &\;\text{for}\; x <0\\
 \end{array}
\right.\\[2ex]
& = 
\sqrt{\frac{2}{\pi}}\;b\; \exp\left(\frac{2 \frac{1}{b}  - \frac{1}{b} }{2b}\right) \quad \;\text{for}\; x\in\mathbb{R},
\end{align*}
$$
which shows that 
$$
\begin{align*}
\frac{f(x)}{g(x|b)}\leq \sqrt{\frac{2}{\pi}} \; b\;\exp\left(\frac{1}{2b^2}\right).
\end{align*}
$$
Now, to show that the minimum of the bound (in $b>0$) is attained at $b=1,$ we minimize 
$$
b\;\exp\left(\frac{1}{2b^2}\right)
$$
with respect to $b>0,$ which yields
$$
\begin{align*}
1\;\exp\left(\frac{1}{2b^2}\right) + b\;\exp\left(\frac{1}{2b^2}\right)\cdot \left(-2\frac{1}{2b^3}\right) &\overset{!}{=} 0\\[2ex]
1\; - 2\frac{1}{2b} &= 0\\[2ex]
b &= 1
\end{align*}
$$
where we used that $\exp\left(\frac{1}{2b^2}\right)>0.$

```{r}
myfun2 <- function(b){b * exp( 1/(2*b^2) )}
bb <- seq(.5, 2, len = 500)
yy <- myfun2(b = bb)
plot(x = bb, y = yy, type = "l")
abline(v = 1, lty=2)
```

1. 
   (a) Deriving the probability of accepting $Y$ from a simulation 
$$
(Y,U)\sim\mathcal{U}\left(\left\{(y,u)|y\in \operatorname{supp}(f_X)\;\text{and}\;0\leq u\leq Mg(x)\right\}\right).
$$ 

$$
\begin{align*}
P\left(\text{Accepting }Y\right)
&=P\left(U\leq f_X(Y)\right),
\end{align*}
$$
where $U\sim\mathcal{U}\left[0,Mg(Y)\right].$ Standardizing the uniform distribution yields
$$
\begin{align*}
P\left(\text{Accepting }Y\right)
&=P\left(\frac{U}{Mg(Y)}\leq \frac{f_X(Y)}{Mg(Y)}\right),\\[2ex]
\end{align*}
$$
where dividing by $Mg(Y)$ is allowed, since $Mg(y)>0$ for all $y\in\operatorname{supp}(f_X).$ Using that 
$$
\frac{U}{Mg(Y)} \sim\mathcal{U}\left[0,1\right],
$$ 
we have
$$
\begin{align*}
P\left(\text{Accepting }Y\right)
&=\int_{-\infty}^{\infty}\int_0^{\frac{f_X(y)}{Mg(y)}} 1 \; du\; g(y) \;dy\\[2ex] 
&=\int_{-\infty}^{\infty}\;\;\;\frac{f_X(y)}{Mg(y)}\;\; g(y) \;\;dy\\[2ex] 
&=\frac{1}{M}\int_{-\infty}^{\infty}\;f_X(y) \;\;dy\\[2ex] 
&=\frac{1}{M}
\end{align*}
$$
<!-- Note that this also is the ratio between the surface $\{\}$ of the subgraph of 
$$
\begin{align*}
\frac{1}{M} = \frac{\int f_X(y)dy}{\int M g(y)dy}
\end{align*}
$$ -->
  (b) 
  $$
  \begin{align*}
  f_X(x) & \leq M g(x)\quad x\in\operatorname{f_X}\\[2ex]
  \int f_X(x) dx & \leq \int M g(x)dx\\[2ex]
  1              & \leq M \cdot 1\\[2ex]
  \end{align*}
  $$






<!-- 1.  If $U_1,\dots,U_n\overset{\text{i.i.d.}}{\sim}\mathcal{U}[0,1],$ then with $\mathbb{E}(U_1)=\mu=1/2$ and $\mathbb{V}(U_1)=\sigma^2=1/12,$
$$
\begin{align*}
Y_n
& =\frac{\sqrt{n}\left(\bar{U}_n - \mu\right)}{\sigma}\\[2ex]
& = \sqrt{12\; n}\left(\bar{U}_n - \frac{1}{2}\right) \\[2ex]
& \to_d \mathcal{N}(0,1).
\end{align*}
$$
  (a) Largest value $Y_n$ can take: 
     * Largest value $U_1$ can take is $1.$ 
     * Largest value $\bar{U}_n$ can take is $\frac{1}{n} n 1 = 1.$ 
     * Largest value $Y_n$ can take is $\sqrt{12\; n}\frac{1}{2}=\sqrt{3\;n}$
     By the same arguments, the smallest value $Y_n$ can take is $\sqrt{3\;n},$ thus the range of $Y_n$ is **bounded** for each given sample size $n,$ 
     $$
     Y_n\in\left[-\sqrt{3\; n}, \sqrt{3\; n}\right],
     $$
     while the standard normal distribution has **unbounded** range from $-\infty$ to $\infty$
  (b) Next, we derive the moments of $Y_n.$ 
     $$
     \begin{align*}
     M_{Y_n}(t) 
     & = \mathbb{E}\left(\exp\left(t Y_n \right)\right) \\[2ex] 
     & = \mathbb{E}\left(\exp\left(t \; \sqrt{12\; n}\left(\bar{U}_n - \frac{1}{2}\right)  \right)\right) \\[2ex] 
     & = \mathbb{E}\left(\exp\left(t \; \sqrt{12\; n}\left(\frac{1}{n}\sum_{i=1}^n U_i - \frac{1}{2}\right)  \right)\right) \\[2ex] 
     & = \mathbb{E}\left(\exp\left(t \; \sqrt{12\; n}\left(\frac{1}{n}\sum_{i=1}^n \left(U_i - \frac{1}{2}\right)\right)  \right)\right) \\[2ex] 
     & = \mathbb{E}\left(\exp\left( \frac{t \;\sqrt{12}}{\sqrt{n}}\sum_{i=1}^n \left(U_i - \frac{1}{2}\right)  \right)\right) \\[2ex] 
     & = \mathbb{E}\left(\exp\left( \sum_{i=1}^n \frac{t \;\sqrt{12}}{\sqrt{n}}\left(U_i - \frac{1}{2}\right)  \right)\right) \\[2ex] 
     & = \prod_{i=1}^n\mathbb{E}\left(\exp\left( \frac{t \;\sqrt{12}}{\sqrt{n}}\left(U_i - \frac{1}{2}\right)  \right)\right) \\[2ex] 
     & = \left(\mathbb{E}\left(\exp\left( \frac{t \;\sqrt{12}}{\sqrt{n}}\left(U_i - \frac{1}{2}\right)  \right)\right)\right)^n \\[2ex] 
     & = \left(\mathbb{E}\left(\exp\left( \frac{t U_i\;\sqrt{12}}{\sqrt{n}} - \frac{1}{2}\frac{t \;\sqrt{12}}{\sqrt{n}}  \right)\right)\right)^n \\[2ex]
     & = \left(\mathbb{E}\left(\exp\left( \frac{t U_i\;\sqrt{12}}{\sqrt{n}} + \frac{ - t \;\sqrt{3}}{\sqrt{n}}  \right)\right)\right)^n \\[2ex] 
     & = \left(\exp\left(\frac{ - t \;\sqrt{3}}{\sqrt{n}}\right)\right)^n
          \left(\mathbb{E}\left(\exp\left( \frac{t U_i\;\sqrt{12}}{\sqrt{n}}   \right)\right)\right)^n \\[2ex]
     & = \left(\exp\left(\frac{ - t \;\sqrt{3}}{\sqrt{n}}\right)\right)^n
          \left(\mathbb{E}\left(\exp\left( \frac{t \sqrt{12}}{\sqrt{n}} U_i   \right)\right)\right)^n \\[2ex] 
     & = \left(\exp\left(\frac{ - t \;\sqrt{3}}{\sqrt{n}}\right)\right)^n
          \left( \frac{\exp\left( \frac{t \sqrt{12}}{\sqrt{n}} -1 \right)}{\frac{t \sqrt{12}}{\sqrt{n}}}\right)^n \\[2ex] 
     & = \left(\exp\left(\frac{ - t \;\sqrt{3}}{\sqrt{n}}\right)\right)^n
          \left( \frac{t \sqrt{12}}{\sqrt{n}} \exp\left( \frac{t \sqrt{12}}{\sqrt{n}} -1 \right)\right)^n \\[2ex] 
     \end{align*}
     $$
     Since 
     $$
     M_U(t) = \mathbb{E}\left(\exp\left(t U\right)\right)=\frac{\exp(t) - 1}{t}.
     $$ 
     we also have  -->