<!-- LTeX: language=en-US -->
# Nonparametric Density Estimation

## Introduction


```{r}
#| echo: false
data_path <- "81b9ff859175053edfc3c2eb024a87b782aa47fa/data/"
data_path <- "data/"
# Libraries
suppressPackageStartupMessages(library("ggplot2"))    # Plotting 
suppressPackageStartupMessages(library("np"))         # Nonparametric Statistics
suppressPackageStartupMessages(library("KernSmooth")) #
suppressPackageStartupMessages(library("gridExtra"))  # To arrange multiple plots 
suppressPackageStartupMessages(library("gtable")) 
suppressPackageStartupMessages(library("scales"))     # Transparent colors 
```


#### **Example: Income Data Analysis** {-}


```{r}
#| echo: false
inc76 <- read.table(paste0(data_path,"inc76.txt"), col.names="income")
```

```{r}
c(inc76$income)[1:6]
```

Typical aim: Characterizing the income distribution (density function) $f$ given a random sample 
$$
\{X_1,\dots,X_n\}
$$
with 
$$
X_i\overset{\text{i.i.d.}}{\sim} f
$$


Traditional statistical key figures such as 

* Empirical mean 
* Empirical median
* Empirical variance 
* Empirical interquartile distance, etc. 

**Problem:** Such key figures only summarize single aspects of a distribution. 

Estimating the total density function $f$ can provide more detailed, more wholistic information. 


#### **Simplest density estimator: Histogram** {-}

Histogramm für  FES Einkommensdaten im Jahr 1976:

```{r}
#| echo: false
#| fig-cap: Histrogram of income data
hist(inc76$income, 
     freq = FALSE,
     xlab = "Income", 
     ylab = "Density", main = "")
```


Disadvantages of the histogram: 

* Choice of the bin width (and of the start point)
* Discontinuous, locally constant $\Rightarrow$ A histogram cannot be a very efficient estimator for a (continuous) density function $f(x).$


**Example:** Too small or too large bin widths

```{r}
#| echo: false
#| fig-cap: Histrograms of income data for different choices of the bin width. 
par(mfrow = c(1,2))
hist(inc76$income, 
     freq = FALSE,
     breaks = 25,
     xlab = "Income", 
     ylab = "Density", main = "")
hist(inc76$income, 
     freq = FALSE,
     breaks = 4,
     xlab = "Income", 
     ylab = "Density", main = "")
par(mfrow = c(1,1))
```


### From the Histogram to Kernel Density Estimation

Comparison: Histogram versus a kernel density estimation 

```{r}
#| echo: false
#| fig-cap: Histogram versus a nonparametric kernel density estimation (green solid line).
hist(inc76$income, 
     freq = FALSE,
     xlab = "Income", 
     ylab = "Density", main = "")
lines(density(inc76$income), col = "darkgreen")
```

Consider a histogram with $J$ (e.g., $J=8$) bins all having the same bin-width $2h,$ defined by equidistant intervals
$$
(x_{j-1},x_j]
$$ 
with 
$$
x_j-x_{j-1}=2h\quad \text{for all} \quad j=1,\dots,J.
$$
The bin-height is determined **locally** at the $j$th interval mid-point 
$$
\bar{x}_j=(x_{j-1}+x_j)/2
$$
by the relative frequency of data points $X_1,\dots,X_n$ that fall within the $j$th interval $(x_{j-1},x_j],$
$$ 
\begin{align*}
\hat f_{hist}(\bar{x}_j)
& =\frac{\hbox{Number of } X_{i}\hbox{ in } (x_{j-1},x_j]}{2hn}\\[2ex]
& =\frac{1}{nh}\sum_{i=1}^n K\left(\frac{X_{i}-\bar{x}_j}{h}\right)
\end{align*}
$$
with
$$
K(z)=\left\{
\begin{array}{ll}
1/2 & \hbox{ if } z\in (-1,1] \\
0   & \hbox{ else}.
\end{array}\right.
$$

Note: The scaling by $2hn$ is necessary to guarantee that the area of each bin equals the relative frquency of data points $X_1,\dots,X_n$ that fall into the interval $(x_{j-1},x_j]$
$$
\begin{align*}
\text{Area of $j$th Bin} 
& = (\text{bin-width}) \cdot (\text{bin-height of $j$th bin})\\[2ex]
& = \qquad 2h\quad\, \cdot \frac{\hbox{Number of } X_{i}\hbox{ in } (x_{j-1},x_j]}{2hn}\\[2ex]
& = \frac{\hbox{Number of } X_{i}\hbox{ in } (x_{j-1},x_j]}{n}%\\[2ex]
\end{align*}
$$
This guarantees that the bin areas of the histogram sum up to one.  


A kernel density estimator generalizes the histogram by estimating the unknown density $f$ at **every** $x$ using 
$$
\begin{align*}
{\hat{f}}_{h}(x)=\frac{1}{nh}\sum_{i=1}^nK\left(
\frac{x-X_{i}}{h}\right)
\end{align*}
$$

* $K$ kernel function 
* $h$ bandwidth 


## Motivating the Kernel Density Estimator


Let 
$$
\{X_1, \ldots, X_n\}
$$
denote a random sample with 
$$
X_i\overset{\text{i.i.d.}}{\sim} f\quad\text{for}\quad i=1,\dots,n.
$$


Aim: Find a density estimator $\hat{f}$ for the true, unknown density $f$ without making a parametric assumption such as assuming that $f$ is the density of a normal distribution with unknown mean and unknown variance. 


**Qualitative assumption on $f:$**

The density function $f(x)$ is sufficiently smooth (i.e., is sufficiently often differentiable) for all $x$.  


Starting point: 

We use the connection between density functions $f$ and distribution functions $F(x) = P(X \leq x),$ 
$$
\begin{equation*}
  f(x) = \frac{d}{dx} F(x) = F'(x), \qquad x \in \mathbb{R}
\end{equation*}
$$

**Idea:** 

Approximate the derivative of the distribution function using the difference quotient. For (small) $h > 0,$ we have 
$$
\begin{align*}
  F'(x) &\approx \frac{F(x+h) - F(x)}{h}\\[2ex]
  F'(x) &= \frac{F(x+h) - F(x)}{h} + O(h)
\end{align*}
$$
or
$$
\begin{align*}
F'(x) &\approx \frac{F(x) - F(x-h)}{h}\\[2ex]
F'(x)      &= \frac{F(x) - F(x-h)}{h} + O(h)
\end{align*}
$$

In both cases, the approximation error is of the order $O(h)$ for $h\to 0,$
$$
\begin{align*}
F'(x) - \frac{F(x) - F(x-h)}{h} = O(h)
\end{align*}
$$
(Pronounce "Big-Oh $h$"). This means that the approximation error goes, in absolute values, to zero as fast as $h\to 0$ or faster; i.e.
$$
\left|F'(x)-\frac{F(x) - F(x-h)}{h}\right|\to 0\quad\text{as}\quad h\to 0 
$$
such that
$$
\frac{\left|F'(x)-\frac{F(x) - F(x-h)}{h}\right|}{h}\to c,
$$
where $0\leq c < \infty.$